Playing with PCAs

### 2024-04-05 3-reports/playing_with_pcas.Rmd
cv and distance

@@

```{r}
cv = 
  data_combined %>% 
  group_by(region, site, transect, horizon, analysis, name) %>% 
  dplyr::summarise(cv = 100 * sd(value)/mean(value),
                   n = n(),
                   mean = mean(value)) %>% 
  filter(!transect %in% "wte")
```

```{r}
(cv %>% 
   filter(!(name == "Al_meq100g" & region == "WLE")) %>% 
  ggplot(aes(x = transect, y = cv))+
  geom_violin(scale = "area", fill = "grey90")+
  geom_jitter(width = 0.1, size = 3, aes(color = site, text = name))+
   geom_boxplot(width = 0.05, outlier.color = NA)+
  facet_wrap(~region)+
   theme_kp())  %>% plotly::ggplotly()

cv %>% 
  ggplot(aes(x = region, y = cv))+
  geom_point(width = 0.1, aes(color = transect), 
             position = position_dodge(width = 0.2))+
  expand_limits(y = 0)+
  facet_wrap(~ name, scales = "free_y")

@@

>>>>

```{r}
wine.pca <- prcomp(wine[treatments], scale = TRUE)

df.wine.x$groups <- wine.class
wine$groups <- wine.class
feature.centroids <- aggregate(wine[treatments], list(Type = wine$groups), mean)

data(wine)
treatments <- 1:2 #treatments to be considerd for this calculation
wine.pca <- prcomp(wine[treatments], scale. = TRUE)
#calculate the centroids for the feature/treatment space and the pca space
df.wine.x <- as.data.frame(wine.pca$x)
df.wine.x$groups <- wine.class
wine$groups <- wine.class
feature.centroids <- aggregate(wine[treatments], list(Type = wine$groups), mean)

dist(rbind(feature.centroids[feature.centroids$Type == "barolo",][-1],
           feature.centroids[feature.centroids$Type == "grignolino",][-1]), method = "euclidean")

data.pca = prcomp(data_wide_PCA %>% drop_na() %>% dplyr::select(where(is.numeric)))
data.feature.centroids <- aggregate(data_wide_PCA %>% drop_na() %>% dplyr::select(where(is.numeric)), 
                                    list(Type = (data_wide_PCA %>% drop_na())$region), 
                                    mean, na.action = na.pass)

dist(rbind(data.feature.centroids[data.feature.centroids$Type == "CB",][-1],
           data.feature.centroids[data.feature.centroids$Type == "WLE",][-1]), method = "euclidean")

centroid.pca = fit_pca_function(data_wide_PCA)$pca_int
centroid.feature = aggregate(data_wide_PCA)

data.feature.centroids = 
  data_combined_subset %>%
  filter(site == "GWI") %>% 
  group_by(transect, name) %>% 
  dplyr::summarise(mean = mean(value, na.rm = T)) %>% 
  pivot_wider(names_from = "name", values_from = "mean")

dist(rbind(data.feature.centroids[data.feature.centroids$transect == "transition",][-1],
           data.feature.centroids[data.feature.centroids$transect == "wetland",][-1]), method = "euclidean")

dist(rbind(data.feature.centroids[data.feature.centroids$transect == "upland",][-1],
           data.feature.centroids[data.feature.centroids$transect == "wetland",][-1]), method = "euclidean")

dist(rbind(data.feature.centroids[data.feature.centroids$transect == "transition",][-1],
           data.feature.centroids[data.feature.centroids$transect == "upland",][-1]), method = "euclidean")

df = data(iris)
dist.iris<-vegdist(iris[,1:4],method = "bray") #calculate Bray-Curtis dissim
dist.iris

#avg distance to group centroids
permdisp<-betadisper(dist.iris,iris$Species,type="centroid")
permdisp
library(usedist)
dist_between_centroids(dist.iris,1:50, 51:100) #1:50 = group setosa, 51:100 = group versicolor

library(vegan)
TukeyHSD(permdisp)

dat.owc = data_wide_PCA %>% filter(site == "OWC") %>% drop_na() 
dist.owc <- vegdist(dat.owc %>% dplyr::select(where(is.numeric)), 
                    method = "bray") #calculate Bray-Curtis dissim
permdisp.owc <- betadisper(dist.owc, dat.owc$transect, type = "centroid")
TukeyHSD(permdisp.owc)

dat.gcw = data_wide_PCA %>% filter(site == "GCW") %>% drop_na() 
dist.gcw <- vegdist(dat.gcw %>% dplyr::select(where(is.numeric)), 
                    method = "bray") #calculate Bray-Curtis dissim
permdisp.gcw <- betadisper(dist.gcw, dat.gcw$transect, type = "centroid")
TukeyHSD(permdisp.gcw)

dat.gwi = data_wide_PCA %>% filter(site == "GWI") %>% drop_na() 
dist.gwi <- vegdist(dat.gwi %>% dplyr::select(where(is.numeric)), 
                    method = "bray") #calculate Bray-Curtis dissim
permdisp.gwi <- betadisper(dist.gwi, dat.gcw$transect, type = "centroid")
TukeyHSD(permdisp.gwi)

```{r}
library(mosaic)
x1 <- c(1,0,0); x2 <- c(1,2,3); y1 <- c(3,4,5); y2 <- rnorm(3)
# projection onto the 1 vector gives the mean vector
mean(y2)            
mosaic::project(y2, 1)
# return the length of the vector, rather than the vector itself
project(y2, 1, type='length')
project(y1 ~ x1 + x2) -> pr; pr

project(x, y, type = "length")

@@

```{r}
cv = 
  data_combined %>% 
  group_by(region, site, transect, horizon, analysis, name) %>% 
  dplyr::summarise(cv = 100 * sd(value)/mean(value),
                   sd = sd(value),
                   n = n(),
                   mean = mean(value)) %>% 
  filter(!transect %in% "wte") %>% 
  filter(!(name == "Al_meq100g" & region == "WLE")) %>% 
  filter(name != "gwc_perc")

```{r}
(cv %>% 
   filter(!(name == "Al_meq100g" & region == "WLE")) %>% 
  ggplot(aes(x = transect, y = cv))+
  geom_violin(scale = "area", fill = "grey90")+
  geom_jitter(width = 0.1, size = 3, aes(color = site, text = name))+
   geom_boxplot(width = 0.05, outlier.color = NA, fill = "white")+
  facet_wrap(~region)+
   theme_kp())  %>% plotly::ggplotly()

cv %>% 
  ggplot(aes(x = region, y = cv))+
  geom_point(width = 0.1, aes(color = transect), 
             position = position_dodge(width = 0.2))+
  expand_limits(y = 0)+
  scale_y_log10()+
  facet_wrap(~ name, 
             #scales = "free_y"
             )

cv %>% 
  filter(!(name == "Al_meq100g" & region == "WLE")) %>% 
  ggplot(aes(x = name, y = cv))+
  geom_violin()+
  geom_point(width = 0.1, aes(color = region), 
            # position = position_dodge(width = 0.9)
             )+
  expand_limits(y = 0)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_grid(transect ~ .)+
#  scale_y_log10()+
  NULL

cv %>% 
  filter(!(name == "Al_meq100g" & region == "WLE")) %>% 
  ggplot(aes(y = name, x = cv))+
  geom_violin(aes(color = region))+
  geom_point(width = 0.1, aes(color = region), 
             position = position_dodge(width = 0.9)
             )+
  expand_limits(x = 0)+
#  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~transect)+
#  scale_y_log10()+
  NULL

cv %>% 
  ggplot(aes(x = mean, y = cv, color = region))+
  geom_point()+
  facet_wrap(~name, scales = "free_x")

@@

```{r}
library(ggwaffle)
cv %>% 
  ggplot(aes(x = transect, y = name, fill = cv))+
  geom_waffle()+
  geom_text(aes(label = round(cv,2)), color = "black")+
  scale_fill_gradientn(trans = "log",
                       colors = PNWColors::pnw_palette("Bay", 12), breaks = c(0, 10, 20, 40, 100, 200))+
  facet_wrap(~region + site)

cv[cv == "NaN"] <- NA
cv = v %>% drop_na()

aov(cv ~ transect + region + name, data = cv) %>% summary()

library(nlme)
nlme::lme(cv ~ transect, random = ~1|name, data = cv %>% drop_na() %>% filter(region == "WLE")) %>% anova()
nlme::lme(cv ~ region, random = ~1|transect, data = cv %>% drop_na()) %>% anova()

cv %>% group_by(region, transect) %>% dplyr::summarise(mean_cv = mean(cv, na.rm = T))

a = aov((cv) ~ transect, data = cv %>% filter(region == "CB"))
agricolae::HSD.test(a, "transect") %>% print()

a = aov((cv) ~ transect, data = cv %>% filter(region == "WLE"))
agricolae::HSD.test(a, "transect") %>% print()
```
